---
title: "Title"
author: "Ishani Makwana, Henry Hirsch and Ei Tanaka"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
rm(list=ls())
library(readr)
library(reshape2)
library(corrplot)
library(dplyr)
library(usmap)
library(ggplot2)
library(gridExtra)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

## Introduction (need edited)
**(need edited)**
```{r init_data}
# Importing a geographic health disparities data set from github repo
urlfile = 'https://raw.githubusercontent.com/eitanaka/DATS6101_Project1_Team2/main/dataset/geographic_health_disparities.csv'

geo_health_df <- read_csv(url(urlfile))
```

## Summary Statistics(need edited)
```{r}
# Create a list containing the data frames by each state
data_by_state <- split(geo_health_df, geo_health_df$StateAbbr)
```

### Depression by state(The variable about Health outcome)
```{r EDA_basicstats_Depression}
# Summary statistic about percentage of population affected with depression in each state
summary_by_state_DEPRESSION <- lapply(data_by_state, function(x) summary(x$Data_Value.DEPRESSION))
head(summary_by_state_DEPRESSION,3)
tail(summary_by_state_DEPRESSION,3)

# A list of standard deviation for percentage of population affected with Depression in each state
sd_by_stateD<- lapply(data_by_state, function(x) sd(x$Data_Value.DEPRESSION))
head(summary_by_stateD,3)
tail(summary_by_stateD,3)

# A data frame containing mean values of the incidence rate of Depression for each country in each state
mean_by_state_DEPRESSION <- lapply(data_by_state, function(x) mean(x$Data_Value.DEPRESSION))
mean_df_DEPRESSION <-data.frame(State = names(mean_by_state_DEPRESSION), Mean_Depression=unlist(mean_by_state_DEPRESSION))

# Add column "fips" into the mean_df_Depression data frame.
mean_df_DEPRESSION["fips"] <- fips(mean_df_DEPRESSION$State)

# A map of the U.S. showing average percent of people suffering from DEPRESSION for each country in each state
plot_usmap(data=mean_df_DEPRESSION, values="Mean_Depression", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "red", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Depression by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

### Mental health by state(The variable about Health Status)
```{r echo=T, eval=T, include=T, results='markup'}
# Summary of the percentage of the population in each country in each state with a health status of 14 or more days with poor mental health
summary_by_state_MHLTH <- lapply(data_by_state, function(x) summary(x$Data_Value.MHLTH))
head(summary_by_state_MHLTH,3)
tail(summary_by_state_MHLTH,3)

# The list containing standard deviation per state: The percentage of the population in each country with a health status of 14 or more day with poor mental health
sd_by_state_MHLTH <- lapply(data_by_state, function(x) sd(x$Data_Value.MHLTH))
head(sd_by_state_MHLTH,3)
tail(sd_by_state_MHLTH,3)

# The data frame containing mean value per state: Percentage of the population in each country with poor mental health for 14 or more days as a health status.
mean_by_state_MHLTH <- lapply(data_by_state, function(x) mean(x$Data_Value.MHLTH))
mean_df_MHLTH <- data.frame(State = names(mean_by_state_MHLTH), Mean_MHLTH = unlist(mean_by_state_MHLTH))

# Add fips column in mean_df_MHLTH
mean_df_MHLTH["fips"] <- fips(mean_df_MHLTH$State)

# Map of the United States plotting the mean percent incidence of the population with poor mental health by state.
plot_usmap(data=mean_df_MHLTH, values="Mean_MHLTH", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "green", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Mental Health by State") +
  guides(fill = guide_colorbar(title = " Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

### Sleep by state(The variable about Health Risk Behabior)
```{r echo=T, eval=T, include=T, results='markup'}
summary_by_stateS <- lapply(data_by_state, function(x) summary(x$Data_Value.SLEEP))
head(summary_by_stateS,3)
tail(summary_by_stateS,3)

sd_by_stateS <- lapply(data_by_state, function(x) sd(x$Data_Value.SLEEP))
head(sd_by_stateS,3)
tail(sd_by_stateS,3)

# Convert mean_by_state to a data frame
mean_by_state_SLEEP <- lapply(data_by_state, function(x) mean(x$Data_Value.SLEEP))
mean_df_SLEEP <- data.frame(State = names(mean_by_state_SLEEP), Mean_Sleep = unlist(mean_by_state_SLEEP))

# # Create a bar chart of the mean depression values by state
# barplot(mean_df_SLEEP$Mean_Sleep, names.arg = mean_df_SLEEP$State, xlab = "State", ylab = "Mean SLEEP Value", col = "blue", main = "Mean SLEEP Values by State")

# Create a new maps object
mean_df_SLEEP["fips"] <- fips(mean_df_SLEEP$State)

plot_usmap(data=mean_df_SLEEP, values="Mean_Sleep", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "purple", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Sleep by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

### LPA by state (A variable about Health Risk Behavior)
```{r}
summary_by_stateS <- lapply(data_by_state, function(x) summary(x$Data_Value.LPA))
head(summary_by_stateS,3)
tail(summary_by_stateS,3)

sd_by_stateS <- lapply(data_by_state, function(x) sd(x$Data_Value.LPA))
head(sd_by_stateS,3)
tail(sd_by_stateS,3)

# Convert mean_by_state to a data frame
mean_by_state_LPA <- lapply(data_by_state, function(x) mean(x$Data_Value.LPA))
mean_df_LPA <- data.frame(State = names(mean_by_state_LPA), Mean_LPA = unlist(mean_by_state_LPA))

# Create a new maps object
mean_df_LPA["fips"] <- fips(mean_df_LPA$State)

plot_usmap(data=mean_df_LPA, values="Mean_LPA", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "cyan", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Lack of Physial Activity by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```


### Corelation between each disease by state
```{r}
# Compute the correlation matrix
# Merge the first two data frames by the "id" column
mean_df <- merge(merge(mean_df_DEPRESSION, mean_df_MHLTH, by=c("State", "fips")), mean_df_SLEEP, by=c("State", "fips"))
mean_df <- merge(mean_df, mean_df_LPA, by=c("State", "fips"))
head(mean_df)
```


### Comparing the mean values of all three variables using boxplot. 

```{r}
# Create a data frame with mean values for each variable


# # Create a box plot
# ggplot(data, aes(x = variable, y = value, fill = variable)) +
#   geom_boxplot() +
# #  facet_wrap(~ variable, ncol=4)
#   geom_jitter(width = 0.2, size = 1, alpha = 0.5) +
#   scale_fill_manual(values = c("blue", "red", "green", "yellow")) +
#   labs(title = "Mean Values of Depression, Sleep, LPA and Mental Health")

# ^ figure out how to change green to yellow
```


This code will create a table showing the mean, median, and standard deviation for each variable (depression, sleep, and mental health) across all states, grouped by state names or FIPS codes. You can compare the results to see how each variable varies across different states.

```{r}
# create mean dataframe for all four conditions
colnames(mean_df)[3:6] <- c("Depression", "MHLTH", "Sleep", "LPA")

# create national correlation matrix
cor_matrix <- cor(mean_df[ , c("Depression", "MHLTH", "Sleep", "LPA")])

# create national correlation heat map
cor_heat_map <- corrplot(cor_matrix, method = "color")
cor_heat_map

# create mixed national correlation heat map
mixed_cor_heat_map <- corrplot.mixed(cor_matrix)
mixed_cor_heat_map

# # create correlation list of matrices for states
# cor_by_state <- lapply(data_by_state, function(state_data) {
#   cor(state_data[c("Data_Value.DEPRESSION", "Data_Value.MHLTH", "Data_Value.SLEEP")])
# })
# 
# # create state correlation matrix
# dep_mhlth.cor <- sapply(cor_by_state, function(state) state[1,2])
# dep_slp.cor <- sapply(cor_by_state, function(state) state[1,3])
# mhlth.slp.cor <- sapply(cor_by_state, function(state) state[2,3])
# 
# state_cor_df <- data.frame(state = mean_df$State, dep_mhlth.cor, dep_slp.cor, mhlth.slp.cor)
# state_cor_df["fips"] <- fips(state_cor_df$state)
# state_cor_df <- select(state_cor_df, "state", "fips", "dep_mhlth.cor", "dep_slp.cor", "mhlth.slp.cor")
```

```{r}
# us correlation maps
# plot_usmap(data=state_cor_df, values="dep_mhlth.cor", labels = TRUE) +
#   scale_fill_continuous(low = "white", high = "orange", guide = FALSE) +
#   scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
#   ggtitle("Correlation Betweeen Depression and Mental Health") +
#   guides(fill = guide_colorbar(title = "Correlation", 
#                                 title.position = "top", 
#                                 title.hjust = 0.5, 
#                                 label.position = "left",
#                                 label.hjust = 0.5))
# 
# plot_usmap(data=state_cor_df, values="dep_slp.cor", labels = TRUE) +
#   scale_fill_continuous(low = "white", high = "purple", guide = FALSE) +
#   scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
#   ggtitle("Correlation Betweeen Depression and Sleep") +
#   guides(fill = guide_colorbar(title = "Correlation", 
#                                 title.position = "top", 
#                                 title.hjust = 0.5, 
#                                 label.position = "left",
#                                 label.hjust = 0.5))
# 
# plot_usmap(data=state_cor_df, values="mhlth.slp.cor", labels = TRUE) +
#   scale_fill_continuous(low = "white", high = "green", guide = FALSE) +
#   scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
#   ggtitle("Correlation Betweeen Mental Health and Sleep") +
#   guides(fill = guide_colorbar(title = "Correlation", 
#                                 title.position = "top", 
#                                 title.hjust = 0.5, 
#                                 label.position = "left",
#                                 label.hjust = 0.5))
```

### Linear Regression Test
```{r}
#Linear regression
# data frame is named 'mean_df'
Linear_reg <- lm(Depression ~ Sleep + LPA, data = mean_df)
summary(Linear_reg)


Linear_reg <- lm(MHLTH ~ Sleep + LPA, data = mean_df)
summary(Linear_reg)

```
 we failed to reject depression and there is a significant test for mental health . that means mental health is effected by sleep and LPA. 

### Scatter plot for 2 health risk behavior and a health outcomes ,and Health Status
```{r}
# create scatter plots
# sleep scatter plot
ggplot(mean_df, aes(x = Sleep)) +
  geom_point(aes(y = Depression, color = "Depression")) +
  geom_point(aes(y = MHLTH, color = "MHLTH")) +
  scale_color_manual(name = "Health Outcomes", values = c("Depression" = "red", "MHLTH" = "green")) +
  labs(x = "% Lacking Sleep", 
       y = "% With Health Outcomes",
       title = "Correlation Between Lack of Sleep and Health Outcomes")


# lpa scatter plot
ggplot(mean_df, aes(x = LPA)) +
  geom_point(aes(y = Depression, color = "Depression")) +
  geom_point(aes(y = MHLTH, color = "MHLTH")) +
  scale_color_manual(name = "Health Outcomes", values = c("Depression" = "red", "MHLTH" = "green")) +
  labs(x = "% Lacking Physical Activity", 
       y = "% With Health Outcomes",
       title = "Correlation Between Lack of Physical Activity and Health Outcomes")
```





