---
title: "Life Style and Mental Health"
author: "Ishani Makwana, Henry Hirsch and Ei Tanaka"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
rm(list=ls())
require(readr)
require(reshape2)
require(corrplot)
require(dplyr)
require(usmap)
require(ggplot2)
require(gridExtra)
require(tidyr)
require(ezids)
require(stargazer)
require(car)
require(coefplot)
require(MASS)
require(scatterplot3d)
require(GGally)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r init_data}
# Importing a geographic health disparities data set from github repo
urlfile = 'https://raw.githubusercontent.com/eitanaka/DATS6101_Project1_Team2/main/dataset/geographic_health_disparities.csv'
geo_health_df <- read_csv(url(urlfile))
```

## Abstruct
**Add paragraphs here**

## 1. Introduction:
**Add paragraphs here**

### a. Dataset
**Add paragraphs here**
**Basic information about a dataset: We will look at the dataset county-level geographic health disparities in the US 2020. This dataset offers model-based census tract-level estimates of the prevalence of 29 health outcomes, preventive services usages, chronic disease-related health risk behaviors, and health statuses as part of the 2020 U.S Census.It covers the entire United States- 50 states and the District of Columbia (DC) at the county, place, census tract, and ZIP Code Tabulation Area levels. At four geographic levels, it uniformly offers information on this huge scale for local locations. The Epidemiology and Surveillance Branch of the Centers for Disease Control and Prevention (CDC), Division of Population Health, provided the estimates.**
**These estimates can be used to identify emerging health problems and to help develop and carry out effective, targeted public health prevention activities. Because the small area model cannot detect effects due to local interventions, users are cautioned against using these estimates for program or policy evaluations. Data sources used to generate these model-based estimates include Behavioral Risk Factor Surveillance System (BRFSS) 2020 or 2019 data, Census Bureau 2010 population data, and American Community Survey 2015–2019 estimates.**

### b. SMART Question
**Add paragraphs here**
**Initial Question1: Is there a significant relationship between the prevalence of depression, poor mental health, lack of sleep, lack of physical activity and geographic location in the US?**
**Initial Question2: Are depression and other variables correlated?/ Is there any correlation between depression and other three variables?**
**Research Question: How lack of sleep and less physical activities is impacting human mental health and how it leads to depression issues we are assuming based on modern lifestyle./To what extent are lack of sleep and lack of physical activity correlated with poor mental health and depression among tracts and states in the US?**


### c. Data cleaning and preparation
**After some analysis and research, we decided to work on some specific data, for which we had to clean our dataset. We created a subset keeping only some variables by reshape dataset from long to wide: There are 10 variables in our new subset: Year, State Abbreviation, County Name, County FIPS, Location Name, Total Population (per tract). Furthermore, filtering only rows with health measure of interest.That is Depression, Mental Health, Lack of Sleep, Lack of Leisure Time Physical Activity (estimated % of tract with condition).**

### d. Independent and dependent variable
**There are four main key variables that we would be considering for this dataset.We got two independent variable that includes Lack of physical exercise and lack of sleep.Dependent variable includes depression and Mental health. Here, we made a logo which easily shows how the independent and dependent variables can have a correlation with one another.Additionally, looking at the symbol, it shows that there is a strong correlation among dependent variables(depression and mental health) and we could even see same kind of correlation among independent variables (lack of sleep and no physical exercise).We will check that by performing some EDA and creating plots to analyse the correlation among all four variables and look at their relationships.**

## 2. EDA
**This Exploratory Data Analysis (EDA) section aims to examine and understand the relationships between various health-related variables, such as depression, mental health, sleep, and lack of physical activity. We begin by performing basic EDA, which involves examining the data types, variable names, and the number of observations. We then check for missing values, outliers, and data distribution, addressing any issues that may impact the analysis. Next, we conduct descriptive statistics to understand the data's central tendency, variability, and spread. A series of visualizations and summary statistics for each variable at the national and state levels follow this. We also investigate correlations between variables to identify potential predictors for modeling. Overall, this comprehensive EDA process helps us better understand the data set, uncover patterns, and detect potential anomalies, laying the groundwork for building accurate and reliable predictive models.** 

### a. Basic EDA
**In the Basic EDA section, we scrutinize the dataset's structure and composition by focusing on data types, variable names, and the number of observations. Subsequently, we address missing values and identify outliers to ensure our data is reliable and robust for further analysis. This crucial step establishes a solid foundation for a deeper exploration of the relationships between health-related variables in the subsequent phases of the EDA process.**

#### Data Structure and composition
```{r echo=T, include=T, eval=T, results='markup', collapse=T}
# Look at the datatypes
str(geo_health_df)

# A number of observation
nrow(geo_health_df)
```
**The dataset comprises 72,337 observations and 12 variables, including numeric and character data types. The variables include information on the year, state abbreviation, state description, county name, county FIPS code, location name, and total population. Additionally, the dataset contains values for depression, leisure-time physical inactivity (LPA), sleep, poor general health (PHLTH), and poor mental health (MHLTH). The data represents various health-related indicators for different locations within the United States, which can be further analyzed to identify trends and relationships among these variables.**
**(Add image to easy to interpret what variables in it)**


#### Check for missing values
```{r echo=T, include=T, eval=T, results='markup', collapse=T}
# check for the missing value
sum(is.na(geo_health_df))
```
**In this dataset, we carefully examined missing values across all variables. Fortunately, our dataset is complete, and there are no missing values. This is a significant advantage, as it ensures the reliability and robustness of our analysis, eliminating the need for imputation or other techniques to address missing data. In addition, a complete dataset enables us to move forward confidently in exploring the relationships between health-related variables.**

#### Check for outliers and data distribution
```{r}
# check for outliers
outlier_Depression  <- outlierKD2(geo_health_df, geo_health_df$Data_Value.DEPRESSION)
outlier_MLHTH  <- outlierKD2(geo_health_df, geo_health_df$Data_Value.MHLTH)
outlier_SLEEP  <- outlierKD2(geo_health_df, geo_health_df$Data_Value.SLEEP)
outlier_LPA  <- outlierKD2(geo_health_df, geo_health_df$Data_Value.LPA)
```
**We assessed the presence of outliers in the health-related variables by analyzing their distributions. Both the original and the data without outliers displayed normal distributions for Depression, MHLTH, and Sleep. In contrast, LPA showed a right-skewed distribution. Comparing the datasets revealed minimal differences between those with and without outliers. Therefore, we chose to analyze the dataset containing outliers, expecting a minimal impact on our findings while enabling a comprehensive understanding of the relationships among the health-related variables.**

### b. Descriptive Statistics 
**In the descriptive statistics section, our objective is to understand the central tendency, variability, and spread of various health-related indicators. We analyze the dataset at national and state levels, focusing on critical aspects of health and well-being. By calculating summary statistics, standard deviations, and mean values for each state, we create visual representations using maps and boxplots. This comprehensive analysis allows us to identify trends, patterns, and regional disparities, paving the way for a deeper exploration of the relationships and underlying factors influencing these health-related variables.**
```{r}
# Create a list containing the data frames by each state
data_by_state <- split(geo_health_df, geo_health_df$StateAbbr)
```

#### National level summary
```{r}
summary_nation_DEPRESSION <- summary(geo_health_df$Data_Value.DEPRESSION)
summary_nation_MHLTH <- summary(geo_health_df$Data_Value.MHLTH)
summary_nation_SLEEP <- summary(geo_health_df$Data_Value.SLEEP)
summary_nation_LPA <- summary(geo_health_df$Data_Value.LPA)
summary_nation_DEPRESSION
summary_nation_MHLTH
summary_nation_SLEEP
summary_nation_LPA
```

#### Depression by state(The variable about Health outcome)
**summary statistics about the percentage of the population affected by depression in each state using the summary function.Using the sd function, we can also get each state's standard deviation. **
```{r EDA_basicstats_Depression}
# Summary statistic about percentage of population affected with depression in each state
summary_by_state_DEPRESSION <- lapply(data_by_state, function(x) summary(x$Data_Value.DEPRESSION))
head(summary_by_state_DEPRESSION,3)
tail(summary_by_state_DEPRESSION,3)

# A list of standard deviation for percentage of population affected with Depression in each state
sd_by_state_DEPRESSION<- lapply(data_by_state, function(x) sd(x$Data_Value.DEPRESSION))
head(sd_by_state_DEPRESSION,3)
tail(sd_by_state_DEPRESSION,3)

# A data frame containing mean values of the incidence rate of Depression for each country in each state
mean_by_state_DEPRESSION <- lapply(data_by_state, function(x) mean(x$Data_Value.DEPRESSION))
mean_df_DEPRESSION <-data.frame(State = names(mean_by_state_DEPRESSION), Mean_Depression=unlist(mean_by_state_DEPRESSION))
mean_df_DEPRESSION["fips"] <- fips(mean_df_DEPRESSION$State)

# A map of the U.S. showing average percent of people suffering from DEPRESSION for each country in each state
plot_usmap(data=mean_df_DEPRESSION, values="Mean_Depression", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "red", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Depression by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

#### Mental health by state(The variable about Health Status)
```{r echo=T, eval=T, include=T, results='markup'}
# Summary of the percentage of the population in each country in each state with a health status of 14 or more days with poor mental health
summary_by_state_MHLTH <- lapply(data_by_state, function(x) summary(x$Data_Value.MHLTH))
head(summary_by_state_MHLTH,3)
tail(summary_by_state_MHLTH,3)

# The list containing standard deviation per state: The percentage of the population in each country with a health status of 14 or more day with poor mental health
sd_by_state_MHLTH <- lapply(data_by_state, function(x) sd(x$Data_Value.MHLTH))
head(sd_by_state_MHLTH,3)
tail(sd_by_state_MHLTH,3)

# The data frame containing mean value per state: Percentage of the population in each country with poor mental health for 14 or more days as a health status.
mean_by_state_MHLTH <- lapply(data_by_state, function(x) mean(x$Data_Value.MHLTH))
mean_df_MHLTH <- data.frame(State = names(mean_by_state_MHLTH), Mean_MHLTH = unlist(mean_by_state_MHLTH))
mean_df_MHLTH["fips"] <- fips(mean_df_MHLTH$State)

# Map of the United States plotting the mean percent incidence of the population with poor mental health by state.
plot_usmap(data=mean_df_MHLTH, values="Mean_MHLTH", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "green", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Mental Health by State") +
  guides(fill = guide_colorbar(title = " Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

#### Sleep by state(The variable about Health Risk Behabior)
```{r echo=T, eval=T, include=T, results='markup'}
# A list of summary statistic per state: Each data is a percentage of sleep disturbance in each country
summary_by_state_SLEEP <- lapply(data_by_state, function(x) summary(x$Data_Value.SLEEP))
head(summary_by_state_SLEEP,3)
tail(summary_by_state_SLEEP,3)

# A list of standard diviation per state: Each data is a percentage of sleep disturbance in each country
sd_by_state_SLEEP <- lapply(data_by_state, function(x) sd(x$Data_Value.SLEEP))
head(sd_by_state_SLEEP,3)
tail(sd_by_state_SLEEP,3)

# A data frame about mean value per state: Each data is a percentage of sleep disturbance in each country
mean_by_state_SLEEP <- lapply(data_by_state, function(x) mean(x$Data_Value.SLEEP))
mean_df_SLEEP <- data.frame(State = names(mean_by_state_SLEEP), Mean_Sleep = unlist(mean_by_state_SLEEP))
mean_df_SLEEP["fips"] <- fips(mean_df_SLEEP$State)

# Map of the United States plotting the mean percent incidence of the population with poor sleep by state.
plot_usmap(data=mean_df_SLEEP, values="Mean_Sleep", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "purple", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Sleep by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

#### LPA by state (A variable about Health Risk Behavior)
```{r}
# A list of summary statistic per state: Data are percent of population of people without leisure time in each country
summary_by_state_LPA <- lapply(data_by_state, function(x) summary(x$Data_Value.LPA))
head(summary_by_state_LPA,3)
tail(summary_by_state_LPA,3)

# A list of standard deviation per state: data are percent of population of people without leisure time in each country
sd_by_state_LPA <- lapply(data_by_state, function(x) sd(x$Data_Value.LPA))
head(sd_by_state_LPA,3)
tail(sd_by_state_LPA,3)

# A data frame about mean value per state: data are percent of population of people without leisure time in each country
mean_by_state_LPA <- lapply(data_by_state, function(x) mean(x$Data_Value.LPA))
mean_df_LPA <- data.frame(State = names(mean_by_state_LPA), Mean_LPA = unlist(mean_by_state_LPA))
mean_df_LPA["fips"] <- fips(mean_df_LPA$State)

# Map of the United States plotting the mean percent incidence of the population without leisure time by state.
plot_usmap(data=mean_df_LPA, values="Mean_LPA", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "cyan", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Mean Lack of Physial Activity by State") +
  guides(fill = guide_colorbar(title = "Mean %", 
                                title.position = "top", 
                                title.hjust = 0.5, 
                                label.position = "left",
                                label.hjust = 0.5))
```

```{r}
# create mean_df for all four conditions
mean_df <- merge(merge(mean_df_DEPRESSION, mean_df_MHLTH, by=c("State", "fips")), mean_df_SLEEP, by=c("State", "fips"))
mean_df <- merge(mean_df, mean_df_LPA, by=c("State", "fips"))
colnames(mean_df)[3:6] <- c("Depression", "MHLTH", "Sleep", "LPA")
head(mean_df)
```

#### Comparing the mean values of all three variables using boxplot. 
```{r}
# create a long format of the data
mean_df_long <- gather(mean_df, key = "Variable", value = "Value", -State, -fips)

# group data by Variable and get the max and min values for each group
max_min_df <- mean_df_long %>% group_by(Variable) %>% 
  slice(which.max(Value), which.min(Value)) %>% ungroup()

## create a box plot for each variable and facet by variable
ggplot(mean_df_long, aes(x = "", y = Value, fill = Variable)) +
  geom_boxplot() +
  geom_jitter(aes(color = Variable), width = 0.2, size = 2) +
  facet_wrap(~Variable, ncol = 4, scales = "fixed") +
  scale_fill_manual(values = c("pink", "cyan", "lightgreen", rgb(200, 162, 200, maxColorValue = 255))) +
  scale_color_manual(values = c("darkred", "darkblue", "darkgreen", "darkorchid")) +
  labs(title = "Health Condition Distributions", x="Health Conditions", y = "% of State Pop. with Condition") +
  geom_text(data = max_min_df, aes(x = 1.25, y = Value, label = State), size = 4, fontface = "bold", hjust = -0.2, color = "black") # add state labels for max and min values
```

This code will create a table showing the mean, median, and standard deviation for each variable (depression, sleep, and mental health) across all states, grouped by state names or FIPS codes. You can compare the results to see how each variable varies across different states.

### c. Correlations(Heat Map)
**Exploring the relationships between variables and identifying any correlations that may exist. This can help to identify potential predictors for modeling.**
**Firstly, we will be checking the correlation between health conditions at national level by investigating the relationship between the four health conditions by generating a mixed correlation matrix of the national-level correlations between our variables.With the cor() function in R, we are calculating a correlation matrix. We are concentrating on four variables: depression, mental health (MHLTH), sleep, and LPA (physical activity) on track level(county - 72337).Then utilizing the corrplot() method to generate a heat map from the correlation matrix. This may be a helpful tool to see how different variables are correlated. Another method to display correlations across variables is to create a mixed correlation heat map using the corrplot.mixed() function. The strongest correlations were between Depression and MHLTH, Sleep and LPA, MHLTH and Sleep, and MHLTH and LPA. In other words, states with high levels of depression are also likely to have high levels of poor mental health, states with high levels of inadequate sleep are also likely to have high levels of inadequate physical activity. And those states with high levels of poor mental health are likely to have high levels of both inadequate sleep and inadequate physical activity. However, there is the least correlation observed between MHLTH and sleep. Overall, using these procedures can help you examine the connections between your data's components/variables and find new predictors for modeling.**


```{r}
# Compute the correlation matrix
# create national correlation matrix
cor_matrix <- cor(mean_df[ , c("Depression", "MHLTH", "Sleep", "LPA")])

# create national correlation heat map
corrplot(cor_matrix, method = "color")

# create mixed national correlation heat map
mixed_cor_heat_map <- corrplot.mixed(cor_matrix, 
                                     main = "Correlation Between Health Conditions (National)",
                                     mar = c(0,0,2,0))
mixed_cor_heat_map

# Create a data frame containing state, FIPs code, cor_MHLTH_sleep, and cor_MHLTH_LPA
cor_by_state_matrix <- lapply(data_by_state, function(state) cor(state[c("Data_Value.DEPRESSION", "Data_Value.MHLTH", "Data_Value.SLEEP", "Data_Value.LPA")]))
cor_by_state_df <- data.frame(
    state = names(cor_by_state_matrix),
    MHLTH_SLEEP = sapply(cor_by_state_matrix, function(state) state[2,3]),
    MHLTH_LPA = sapply(cor_by_state_matrix, function(state) state[2,4])
)
cor_by_state_df["fips"] <- fips(cor_by_state_df$state)
```

#### Scatter plot for 2 health risk behavior and a health outcomes ,and Health Status
```{r}
# create scatter plots
# sleep scatter plot
ggplot(mean_df, aes(x = Sleep)) +
  geom_point(aes(y = Depression, color = "Depression")) +
  geom_point(aes(y = MHLTH, color = "MHLTH")) +
  scale_color_manual(name = "Health Outcomes", values = c("Depression" = "red", "MHLTH" = "green")) +
  labs(x = "% Lacking Sleep", 
       y = "% With Health Outcomes",
       title = "Correlation Between Lack of Sleep and Health Outcomes") +
  geom_smooth(aes(y = Depression, color='black'), method = "lm", se = TRUE) +
  geom_smooth(aes(y = MHLTH, color='black'), method = "lm", se = TRUE)


# lpa scatter plot
ggplot(mean_df, aes(x = LPA)) +
  geom_point(aes(y = Depression, color = "Depression")) +
  geom_point(aes(y = MHLTH, color = "MHLTH")) +
  scale_color_manual(name = "Health Outcomes", values = c("Depression" = "red", "MHLTH" = "green")) +
  labs(x = "% Lacking Physical Activity", 
       y = "% With Health Outcomes",
       title = "Correlation Between Lack of Physical Activity and Health Outcomes") +
  geom_smooth(aes(y = Depression, color='black'), method = "lm", se = TRUE) +
  geom_smooth(aes(y = MHLTH, color='black'), method = "lm", se = TRUE)
```

#### Correlation between MHLTH and sleep, MHLTH and LPA using us map.
```{r}
# The map of US about correlation between MHLTH and Sleep.
plot_usmap(data=cor_by_state_df, values="MHLTH_SLEEP", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "blue", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Correlation Betweeen Mental Health and SLEEP") +
  guides(fill = guide_colorbar(title = "Correlation",
                                title.position = "top",
                                title.hjust = 0.5,
                                label.position = "left",
                                label.hjust = 0.5))

# The map of US about correlation between MHLTH and LPA
plot_usmap(data=cor_by_state_df, values="MHLTH_LPA", labels = TRUE) +
  scale_fill_continuous(low = "white", high = "blue", guide = FALSE) +
  scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
  ggtitle("Correlation Betweeen Mental Health and Sleep") +
  guides(fill = guide_colorbar(title = "Correlation",
                                title.position = "top",
                                title.hjust = 0.5,
                                label.position = "left",
                                label.hjust = 0.5))
```

### d. Hypothesis testing
**If applicable, conduct statistical hypothesis tests to confirm or reject hypotheses about the data.**

#### Linear Regression Test
```{r}
# multiple linear regressions

depression.mlg <- lm(Depression ~ Sleep + LPA -1, data = mean_df)
mhlth.mlg <- lm(MHLTH ~ Sleep + LPA -1, data = mean_df)


# mlg tables

depression.mlg.table <- stargazer(depression.mlg,
                                  type = "text",
                                  title = "Multiple Linear Regression for Depression, Sleep, and LPA",
                                  header = FALSE,
                                  digits = 4,
                                  star.cutoffs = c(0.05, 0.01, 0.001),
                                  report = "vcstp*")

mhlth.mlg.table <- stargazer(mhlth.mlg,
                                  type = "text",
                                  title = "Multiple Linear Regression for Mental Health, Sleep, and LPA",
                                  header = FALSE,
                                  digits = 4,
                                  star.cutoffs = c(0.05, 0.01, 0.001),
                                  report = "vcstp*")


# # mlg partial regression plots
# mhlth.lpa.mlg.partial.reg.plot <- qqp(mhlth.mlg, "LPA")
# 
# mhlth.sleep.mlg.partial.reg.plot <- qqp(mhlth.mlg, "Sleep")
# 
# # mlg coefficient plots
# dep.mlg.coefplot <- coefplot(depression.mlg, exclude = TRUE, title = "Depression")
# dep.mlg.coefplot
# 
# mhlth.mlg.coefplot <-coefplot(mhlth.mlg, exclude = TRUE, title = "MHlTH")
# mhlth.mlg.coefplot
# 
# # mlg residual plots
# dep.residual.plot <- plot(depression.mlg, which = 1)
# dep.residual.plot
# 
# mhlth.residual.plot <- plot(mhlth.mlg, which = 1)
# mhlth.residual.plot


```

We failed to reject depression and there is a significant test for mental health . that means mental health is effected by sleep and LPA. 

## 3. Result
**Add paragraphs here**

## 4. Discussion
**Add paragraphs here**
### a. Limitations
**Add paragraphs here**
### b. Future Enforcement
**Add paragraphs here**

## 5. Conclusion
**Add paragraphs here**

## 6. References
**Add references here**

